\documentclass[cameraready]{Interspeech}
\usepackage{float}

\title{Generative Models for Cat Image Synthesis}

\author[affiliation={1}]{Pawel}{Pozorski}
\author[affiliation={1}]{Pawel}{Florek}

\address{
    $^1$ Warsaw University of Technology, Poland
}

\email{\{pawel.pozorski, pawel.florek\}.stud@pw.edu.pl}

\keywords{generative models, GAN, VAE, diffusion, image synthesis, mode collapse, latent interpolation}

\begin{document}

\maketitle

\begin{abstract}
We present a systematic comparison of three generative paradigms---generative adversarial networks (GANs), variational autoencoders (VAEs), and denoising diffusion probabilistic models---for unconditional synthesis of cat images. The study emphasizes experimental reproducibility through fixed configuration management, deterministic pseudorandom seeding, and multi-run variance characterization. A single software stack supports training with checkpoint resumption, experiment logging (MLflow), and a shared evaluation protocol comprising Fr\'echet Inception Distance (FID), perceptual diversity measures, and latent-space interpolation. Beyond the primary cat-only benchmark, we outline an exploratory extension to class-conditional generation on a cats-versus-dogs corpus. This document specifies methodology, situates the work relative to prior art, and formalizes the experimental design; quantitative and qualitative results are deferred pending completion of the full training matrix.
\end{abstract}

\section{Introduction}
Generative modeling seeks to approximate high-dimensional data distributions such that novel samples are statistically faithful to the training domain while exhibiting plausible structure. In image synthesis, three families have become dominant: adversarial training~\cite{goodfellow2014generative}, amortized variational inference with continuous or discrete latents~\cite{kingma2014autoencoding,vandenoord2017neural}, and iterative denoising diffusion~\cite{ho2020denoising}. We investigate all three in a unified setting---synthesis of cat imagery at moderate resolution---so that differences in sample quality, diversity, optimization stability, and sensitivity to initialization may be attributed primarily to the modeling paradigm rather than to unrelated preprocessing or infrastructure choices.

The experimental setting is deliberately bounded by a fixed local setup, which keeps the training and evaluation protocol consistent across runs. Within these limits, we nevertheless adhere to standard quantitative reporting (FID, complementary perceptual metrics) and to qualitative diagnostics of known failure modes, notably mode collapse in adversarial training and oversmoothing in likelihood-based models.

For each family we instantiate a baseline and a full enhanced configuration informed by the literature (e.g., self-attention and spectral normalization in GANs~\cite{zhang2019self,miyato2018spectral}, vector quantization in VAEs~\cite{vandenoord2017neural}, refined schedules and sampling in diffusion~\cite{nichol2021improved}). A separate hyperparameter sweep systematically varies model-specific parameters on both variants to quantify the influence of individual design choices. All primary comparisons are repeated across multiple independent seeds so that point estimates of performance are accompanied by dispersion, reflecting the stochastic nature of non-convex optimization in deep generative modeling.

\subsection{Objectives}
The study is organized around the following objectives:
\begin{enumerate}
    \item to synthesize cat images under each of the three paradigms with matched preprocessing and comparable training budgets;
    \item to contrast baseline and full enhanced variants \emph{within} each paradigm, and to investigate the influence of model-specific hyperparameters via a dedicated sweep;
    \item to quantify distributional fidelity via FID~\cite{heusel2017gans} and to characterize diversity using perceptual and structural similarity measures~\cite{zhang2018unreasonable,wang2003multiscale};
    \item to quantify sensitivity of final metrics to random initialization and training stochasticity across seeds;
    \item to construct latent interpolation trajectories for qualitative assessment of the learned representation;
    \item to conduct a secondary, exploratory analysis of class-conditional generation on a cats-versus-dogs dataset.
\end{enumerate}

\subsection{Contributions}
The principal contributions of this work are as follows:
\begin{itemize}
    \item a modular, strongly typed training and evaluation codebase that treats GAN, VAE, and diffusion models behind a common trainer interface, reducing implementation confounds when comparing variants;
    \item model configurations chosen to remain practical within the shared training setup without abandoning established architectural ideas from the literature;
    \item automated orchestration of a multi-seed, multi-variant experiment matrix with consistent logging and checkpoint semantics;
    \item integration of early stopping and resumable training, which is particularly relevant for interrupted runs;
    \item an optional class-conditional GAN extension for cross-species exploratory synthesis, clearly delineated from the primary cat-only benchmark.
\end{itemize}

\section{Methodology}
\subsection{Datasets}
Two corpora are employed (see Sec.~\ref{sec:exp-design} for acquisition protocol):
\begin{itemize}
    \item a large-scale cat image collection for unconditional synthesis;
    \item a cats-versus-dogs collection for the conditional extension.
\end{itemize}

Image loading is mediated by dataset-specific PyTorch \texttt{Dataset} implementations. Worker processes receive deterministic seeds to improve reproducibility under multi-process data loading. For the cat corpus, the pipeline prefers a project-local directory tree; if that directory contains no usable images and remote retrieval is permitted, archives are obtained via the Kaggle API and materialized into the same root to preserve stable relative paths for analysis. For the extension corpus, class labels are derived heuristically from file naming conventions or subdirectory structure (e.g., prefixes \texttt{cat.} / \texttt{dog.}).

All inputs are resized to a fixed spatial resolution and affine-normalized to $[-1, 1]$, which is the conventional dynamic range for GAN and diffusion training and allows parameter sharing of convolutional front ends across model families where applicable.

\subsection{Model families and variants}
The primary training matrix trains two variants per family---a \emph{baseline} and a \emph{full novel}---each repeated across three seeds. A separate hyperparameter sweep (Sec.~\ref{sec:hparam}) then investigates the influence of model-specific parameters on both variants.

\subsubsection{GAN family}
The baseline is a vanilla DCGAN~\cite{radford2016unsupervised} and the full novel variant is SA-DCGAN+, which adds self-attention~\cite{zhang2019self}, spectral normalization~\cite{miyato2018spectral}, DiffAugment, hinge-style objectives, and R1 gradient regularization. The generator maps a low-dimensional Gaussian latent vector to $64\times 64$ RGB imagery; the discriminator maps the image to a univariate realism score. For the cats-versus-dogs extension, both networks accept class embeddings, realizing a conditional GAN in the sense of~\cite{goodfellow2014generative}.

\subsubsection{VAE family}
The baseline is a Gaussian-latent VAE trained with the ELBO~\cite{kingma2014autoencoding}. The full novel variant is VQ-VAE-P: a VQ-VAE~\cite{vandenoord2017neural} with EMA codebook updates, perceptual reconstruction loss, dead-code reset, residual encoder/decoder blocks, and a gated PixelCNN prior.

\subsubsection{Diffusion family}
The baseline is a vanilla DDPM with a linear variance schedule~\cite{ho2020denoising}. The full novel variant is Improved DDPM+ with a cosine schedule, learned variance, DDIM sampling, and self-conditioning~\cite{nichol2021improved}. A compact time-conditioned U-Net predicts the noise $\epsilon$ at each diffusion step. During training, timesteps are drawn uniformly at random and a denoising objective is minimized in expectation; sampling reverses the forward diffusion chain iteratively.

\subsection{Training framework}
A shared abstract trainer coordinates family-specific inner loops while exposing a uniform interface for logging, checkpointing, and early stopping. Pseudorandom number generators for Python, NumPy, and PyTorch are initialized from a run-level seed. Scalar summaries (losses, monitored validation proxies where defined) are recorded per epoch to MLflow. Checkpoints serialize model parameters, optimizer state, and training step indices to support seamless resumption. Early termination is enabled by default according to configuration: a monitored scalar is tracked with patience and minimum-improvement thresholds, mitigating wasted compute when optimization stagnates or diverges. Higher-level scripts enumerate the Cartesian product of model family, variant label, and seed without manual book-keeping.

\subsection{Experimental structure}
Primary training follows a two-variant-per-family design with shared seed replication. The cat-only benchmark trains a baseline and a full novel variant for each of the three families, each repeated with three seeds ($\{42,123,7\}$), yielding $2\times 3\times 3=18$ independent runs. This structure supports cross-family comparison and baseline-versus-enhanced attribution.

\subsection{Hyperparameter influence study}
\label{sec:hparam}
To satisfy the requirement of investigating the influence of hyperparameters, we conduct a dedicated two-phase sweep per family, focused on \emph{model-specific} parameters (not learning rate or batch size):

\textbf{Phase~1 (baseline sweep):} Each baseline is trained over a grid of its key parameters and three seeds; the configuration yielding the lowest mean final loss is selected as the ``best baseline.''

\textbf{Phase~2 (full-novel sweep):} The best baseline parameters are carried forward as defaults, and the full novel variant is trained over its own grid, again with three seeds per configuration.

The grids are:
\begin{itemize}
    \item \textbf{GAN baseline} (3 configs): \texttt{latent\_dim} $\in\{64,128,256\}$.
    \item \textbf{SA-DCGAN+ full} (6 configs): \texttt{r1\_gamma} $\in\{1,10,50\}$ $\times$ \texttt{r1\_interval} $\in\{8,16\}$.
    \item \textbf{VAE baseline} (6 configs): \texttt{latent\_dim} $\in\{64,128,256\}$ $\times$ \texttt{kl\_weight} $\in\{0.5,2.0\}$.
    \item \textbf{VQ-VAE-P full} (8 configs): \texttt{num\_embeddings} $\in\{256,512\}$ $\times$ \texttt{embedding\_dim} $\in\{64,128\}$ $\times$ \texttt{ema\_decay} $\in\{0.95,0.99\}$.
    \item \textbf{DDPM baseline} (4 configs): \texttt{timesteps} $\in\{500,1000\}$ $\times$ \texttt{sample\_steps} $\in\{250,500\}$.
    \item \textbf{Improved DDPM+ full} (8 configs): \texttt{ddim\_steps} $\in\{50,100\}$ $\times$ \texttt{ddim\_eta} $\in\{0.0,0.2\}$ $\times$ \texttt{self\_cond\_prob} $\in\{0.25,0.5\}$.
\end{itemize}
This yields 35 configurations $\times$ 3 seeds $=$ 105 sweep runs. Combined with the 18 primary matrix runs and the cats-versus-dogs extension, the total experimental budget remains feasible on our hardware.

\subsection{Evaluation protocol}
Evaluation proceeds along four complementary axes:
\begin{itemize}
    \item \textbf{Global fidelity:} FID~\cite{heusel2017gans} between real and generated feature distributions extracted by a pretrained Inception network;
    \item \textbf{Diversity and perceptual spread:} pairwise LPIPS~\cite{zhang2018unreasonable} and MS-SSIM~\cite{wang2003multiscale} statistics on generated batches;
    \item \textbf{Geometric structure of latent space:} piecewise-linear interpolation in the generator or VAE latent space (and analogous trajectories for diffusion as implemented);
    \item \textbf{Robustness:} dispersion of the above metrics across seeds.
\end{itemize}

Dedicated modules implement metric computation; scripts emit rasterized grids and interpolation strips for inclusion in the final manuscript.

\subsection{Cats-versus-dogs extension}
The extension instantiates a class-conditional GAN: class embeddings are concatenated with latent codes in the generator and channel-wise with label maps in the discriminator. Training on imbalanced or mixed-domain imagery is treated as \emph{exploratory}; we emphasize qualitative assessment of class separation versus attribute entanglement rather than leaderboard-style competition scores.

\section{Related Work and Benchmark Context}
\label{sec:related}
\textbf{Adversarial generation.} GANs~\cite{goodfellow2014generative} formulate learning as a two-player game between a generator and a discriminator. DCGAN~\cite{radford2016unsupervised} demonstrated that careful architectural constraints yield stable synthesis of natural images. Nonetheless, adversarial training remains sensitive to hyperparameters and susceptible to mode collapse. Spectral normalization~\cite{miyato2018spectral} constrains the Lipschitz constant of the discriminator, while self-attention~\cite{zhang2019self} captures long-range spatial dependencies; both appear in our full SA-DCGAN+ configuration, and the hyperparameter sweep quantifies the effect of regularization strength.

\textbf{Latent-variable and quantized representations.} VAEs~\cite{kingma2014autoencoding} optimize a tractable lower bound on the marginal log-likelihood with a Gaussian latent prior, often producing slightly blurred reconstructions. VQ-VAE~\cite{vandenoord2017neural} replaces continuous latents with discrete codes drawn from a learned codebook, improving fidelity and enabling powerful autoregressive priors. Our study contrasts the Gaussian baseline against the full VQ-VAE-P variant and sweeps codebook and loss parameters.

\textbf{Diffusion modeling.} DDPMs~\cite{ho2020denoising} define a forward noising process and learn its reverse as a sequence of denoising steps. Subsequent work refined variance schedules and parameterizations to improve sample quality and efficiency~\cite{nichol2021improved}. Diffusion typically demands more function evaluations per sample than single-step generators but offers comparatively stable optimization.

\textbf{Evaluation methodology.} FID~\cite{heusel2017gans} remains a de facto standard for comparing generative models on natural images. LPIPS~\cite{zhang2018unreasonable} measures perceptual distance using deep features; MS-SSIM~\cite{wang2003multiscale} quantifies multiscale structural agreement. Jointly, these metrics provide complementary views on fidelity and intra-batch diversity.

\textbf{Positioning.} Although Interspeech primarily addresses speech and spoken-language processing, the present manuscript applies the same experimental hygiene expected in rigorous machine-learning research---explicit seeds, versioned configurations, and quantitative metrics---to a vision benchmark chosen for visual clarity of failure modes and ease of qualitative inspection.

\section{Experimental Design and Reporting}
\label{sec:exp-design}
\textbf{Compute environment.} Experiments are executed on Apple Silicon with PyTorch compiled for MPS acceleration; the codebase remains portable to CUDA when available. Hyperparameters and architectural constants are declared in a hierarchical YAML configuration to avoid implicit defaults that differ across runs.

\textbf{Primary training matrix.} The cat-only study comprises eighteen training jobs: one baseline and one full novel variant per family, each repeated across three pseudorandom seeds. Each job produces time-series logs, serialized checkpoints, and MLflow metadata sufficient to reconstruct the exact command-line and configuration merge.

\textbf{Hyperparameter sweep.} In addition to the primary matrix, a two-phase sweep of 35 model-specific configurations $\times$ 3 seeds $=$ 105 runs investigates the influence of key hyperparameters (see Sec.~\ref{sec:hparam}). The sweep first selects the best baseline configuration per family, then carries those settings forward to the full novel variant.

\textbf{Optimization and early stopping.} Within each family, optimization hyperparameters are held constant across variants unless that parameter is the explicit subject of the sweep. Early stopping monitors a family-appropriate scalar with fixed patience and minimum relative improvement, terminating training when progress plateaus while still permitting long horizons when monotonic improvement persists. Checkpointing supports recovery from preemption, which is common in mobile workstation settings.

\textbf{Data management.} Cat images reside under a single repository-relative root. If that root is initially empty and downloads are authorized, the acquisition pipeline populates it deterministically from remote sources, avoiding reliance on opaque cache paths for downstream exploratory analysis. The cats-versus-dogs corpus uses a disjoint root with filename-derived labels; results from this stream are explicitly labeled exploratory and are not claimed to match state-of-the-art on the original competition metric.

\textbf{Post-training analysis.} Upon matrix completion, we estimate FID~\cite{heusel2017gans} from a fixed draw of generated samples per checkpoint, summarize LPIPS and MS-SSIM~\cite{zhang2018unreasonable,wang2003multiscale} distributions, and render latent interpolations. Aggregate tables report central tendency and variability across seeds (mean and standard deviation, or median and interquartile range where distributions appear skewed).

\textbf{Upcoming results section.} Section~\ref{sec:results} will consolidate tabular comparisons, seed-sensitivity visualizations, and qualitative panels, linking observed artifacts (e.g., collapse, texture loss, color inconsistency) to the logged quantitative scores and to specific configuration choices.

\section{Results}
\label{sec:results}
\textit{Quantitative metrics, statistical summaries across seeds, hyperparameter influence plots, and qualitative figures will be inserted after the full training matrix and sweep are finalized.}

\section{Conclusion}
\textit{A discussion interpreting the empirical findings in light of Sec.~\ref{sec:related}, together with limitations of the experimental setup and directions for future work, will be added once results are available.}

\section{AI Disclosure}
This manuscript was drafted with assistance from generative language models for phrasing and organization. All scientific judgments, implementation decisions, and interpretations remain the authors' own and theirs responsibility.

\bibliographystyle{IEEEtran}
\bibliography{mybib}

\newpage
~\vfill
\newpage
\appendix
\section{Appendix}
\textit{Supplementary material---extended plots, per-seed learning curves, and implementation notes---will be appended following the main experimental campaign.}

\end{document}
